---
title: "Modulating pH in an Aquaculture Microbiome"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aqua}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Data and Problem Context

```{r, echo = FALSE}
library(RefManageR)
bib <- ReadBib("references.bib")
```

Sometimes we are interested in modeling how an ecosystem shifts under a
continuously modulated input, rather than an abrupt shock This vignette explores
how transfer function models can be adapted to this setting. We re-analyze data from `r Citep(bib, "Yajima2022CoreSA,")`, which studied the microbiome from associated with an eel aquaculture.
Eels were grown in five different tanks, and the microbiomes in each tank were
sequenced daily for 120 days. The study gathered a variety of tank-level
environmental characteristics, including oxygenation and pH. A noticeable shift
in pH occurred for several of the tanks midway through the study. We'll quantify
this effect and use transfer functions to simulate hypothetical trajectories
under continuously-modulated pH changes.

First, let's load the packages used in this vignette. The block defining `th`
just sets `ggplot2` color themes.

```{r setup}
library(mbtransfer)
library(tidyverse)
library(glue)
th <- theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 12),
    legend.position = "bottom"
  )
theme_set(th)
```

We've hosted [lightly preprocessed](https://github.com/krisrs1128/microbiome_interventions/blob/main/scripts/data_preprocessing.Rmd) versions of the [raw data](https://github.com/hiro-toju/EelMicrobiome128) at the following
[figshare links](https://figshare.com/articles/dataset/Processed_Aquaculture_Data/22687216). We'll use the `pH` column in `interventions` as input $w_{nt}$ in a transfer function.

```{r}
reads <- read_csv("https://figshare.com/ndownloader/files/40357894/reads.csv") |>
  column_to_rownames("sample")
samples <- read_csv("https://figshare.com/ndownloader/files/40357897/samples.csv")
subject <- read_csv("https://figshare.com/ndownloader/files/40357891/subject.csv")
interventions <- read_csv("https://figshare.com/ndownloader/files/40357888/interventions.csv") |>
  column_to_rownames("sample")
```

Based on these "long" `data.frames`, we can create a structured `ts` object.
These have "wide," sample-level matrices that are convenient for autoregressive
modeling. There are some missing timepoints in the original data, and we
linearly interpolate these. Finally, from the original publication, three of the
tanks seem to have noticeably different behavior than the rest. They also have a
clear shift in pH in days 40 - 50, so we'll concentrate the remainder of the
analysis on just these tanks.

```{r}
ts <- reads |>
  ts_from_dfs(interventions, samples, subject) |>
  interpolate()

ts <- ts[1:3]
```

## Prediction

We'll evaluate the forecasting performance of a model trained on these three
tanks^[Since we only have three tanks, we won't attempt to generalize to new tanks. There doesn't seem to be enough basis for any kind of tank-level generalization.]. In the block below, we train a model using the first 96 (out
of 128) days within each tank. Then, we forecast the remaining 32 days.

```{r}
ts_missing <- subset_values(ts, 1:96)
fit <- mbtransfer(ts, P = 3, Q = 3)
ts_pred <- predict(fit, ts_missing)
```

We can evaluate performance on these held out days by plotting the truth against
the forecast. Performance is generally better in the more abundant species, and
short time horizons $h$ are easier to predict than long ones.

```{r, fig.width = 10, fig.height = 6}
reshape_preds(ts, ts_pred) |>
  mutate(h_group = cut(h, quantile(h, 5))) |>
  ggplot() +
  geom_abline(slope = 1, col = "#400610") +
  geom_point(aes(y, y_hat), size = .7, alpha = .6) +
  facet_grid(factor(quantile, rev(levels(quantile))) ~ glue("lag {h_group}"), scales = "free_y") +
  labs(x = expression(y), y = expression(hat(y))) +
  scale_x_continuous(expand = c(0, 0), n.breaks = 3) +
  scale_y_continuous(expand = c(0, 0), n.breaks = 3) +
  theme(
    axis.text = element_text(size = 10),
    panel.spacing = unit(0, "line"),
    strip.text.y = element_text(angle = 0, size = 12),
    strip.text.x = element_text(angle = 0, size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 11),
  )
```

### Attribution Analysis: Selecting Important Taxa

Which taxa are most sensitive to changes in pH? Since we trained our transfer
model using pH as an input, we can simulate new trajectories under hypothetical
pH shifts. First, we create a counterfactual with only two pH values -- one
acidic one (2) and another basic one (9). This is just a device for focusing
attention on the taxa that seem to be affected by pH -- we'll later simulate
from a continuous gradient of pH values. Note that our counterfactuals `ws` have
to specify all the perturbations, even though we're really only interested in
one of them. We've set `AS` to 38 because it is a typical value of that series
across all the observed data.

```{r}
ws <- steps(c("pH" = TRUE, "AS" = FALSE), lengths = 4, L = 4) |>
  map(~ 7 * . + 2)
ws[[1]]["AS", ] <- 38
ws[[2]]["AS", ] <- 38

staxa <- select_taxa(ts, ws[[1]], ws[[2]], \(x) mbtransfer(x, 3, 3))
#save(staxa, file = "../../data/staxa-aqua.rds")

taxa <- staxa$ms |>
  select(multisplit, m, lag) %>%
  split(.$lag) %>%
  map(~ split(., .$multisplit) %>% map(~ pull(., m))) |>
  map(~ which(multiple_data_splitting(., q = 0.2)))
```

Let's check the mirror statistics associated with these taxa. We're showing only
the 50 taxa with the largest mirror statistics, but even among these, there are
most don't seem to be affected by pH strongly enough to be flagged at a $q =
0.2$ false discovery rate control level (the default used in mbtransfer, which
can be modified in the `select_taxa` function).

```{r, fig.height = 5, fig.width = 10}
vis_taxa <- staxa$ms |>
  group_by(taxon) |>
  summarise(m = mean(m)) |>
  slice_max(m, n = 50) |>
  pull(taxon)

staxa$ms |>
  filter(taxon %in% vis_taxa) |>
  mutate(
    taxon = taxa(ts)[taxon],
    lag = as.factor(lag),
    selected = ifelse(taxon %in% unlist(staxa$taxa), "Selected", "Unselected")
  ) |>
  ggplot() +
  geom_hline(yintercept = 0, linewidth = 2, col = "#d3d3d3") +
  geom_boxplot(aes(reorder(taxon, -m), m, fill = lag, col = lag), alpha = 0.8) +
  facet_grid(. ~ selected, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = c("#c6dbef", "#6baed6", "#2171b5", "#084594")) +
  scale_color_manual(values = c("#c6dbef", "#6baed6", "#2171b5", "#084594")) +
  labs(y = expression(M[j]), x = "Taxon") +
  theme(axis.text.x = element_text(angle = 90, size = 11))
```

How plausible are these selections? One simple check is to compare each taxon's
abundance against pH across the time series. We've sorted taxa according to the
strength of the association according to the mirror statistic. Note that, though
we could always compute simple $R^2$ from these scatterplots, any test
statistics derived from them would be suspect, since the samples are correlated
over time. The effective sample size is not as large as the number of points in
each plot might lead you to believe. A transfer function + mirror testing
approach accounts for temporal dependence, which is the main reason for going
through all the additional modeling effort.

```{r}
focus_taxa <- unlist(map(staxa$taxa, ~ c(.)))

taxa_order <- staxa$ms |>
  group_by(taxon) |>
  summarise(median_m = median(m)) |>
  arrange(-median_m) |>
  pull(taxon)

pivot_ts(ts) |>
  mutate(taxon = factor(taxon, levels = taxa(ts)[taxa_order])) |>
  filter(taxon %in% focus_taxa, taxon %in% taxa(ts)[vis_taxa]) |>
  ggplot() +
  geom_point(aes(pH, value), alpha = 0.6, size = 0.8) +
  facet_wrap(~ taxon, ncol = 7)
```

Next, we'll see how these taxa vary in response to intermediate shifts pH, not
just the acidic and basic extremes considered above. For this, we need a richer
collection of counterfactuals. The `ws` below vary from 2 to 9, and we assume
that the shift lasts for 42 time steps.

```{r}
L <- 42
ws <- list()
plateaus <- seq(2, 9, length.out = 10)
for (p in seq_along(plateaus)) {
  ws[[p]] <- matrix(38, nrow = 2, ncol = L)
  colnames(ws[[p]]) <- glue("t{seq_len(L)}")
  rownames(ws[[p]]) <- c("pH", "AS")
  ws[[p]][1, ] <- rep(plateaus[p], L)
}
```

The `predict` method for `mbtransfer` models looks at the difference between
`interventions` and `values` slots, which contain the environmental input and
microbial response variables. Timepoints which are present in `interventions`
but not `values` will be used to forecast new `values`. With this in mind, we
can create a dataset that will allow us to simulate trajectories under each of
the `ws`. The idea is to replace the interventions starting at time `L` with our
counterfactual `ws` and then remove all the actually observed values following
that timepoint. Both of these steps are encapsulated in the `counterfactual_ts`
function below. We usually use this function to compare two pairs of
counterfactuals. Since we are actually generating more than two, we'll loop over
this function and place a dummy for the third argument (this would usually
define a second, `ts2` time counterfactual time series).

```{r}
sim_ts_ <- list()
for (i in seq_along(ws)) {
  sim_ts_[[i]] <- counterfactual_ts(ts, ws[[i]], ws[[i]], start_ix = L)$ts1
}
```

Finally, we predict a series under each of the hypothetical pH interventions.
We've visualized a few of the taxa whose trajectories are shifted by different
degrees according to the strength of the intervention. The effects seems
essentially additive. In principle, the response "shapes" could change for
different pH, the result of potential feedback loops between taxa. However, this
does not seem to be the case for these data.

```{r}
pal <- c("#cbd8e8", "#becbdc", "#b1bed0", "#a4b2c4", "#97a5b8", "#8b99ac", "#808da0", "#748193", "#697587", "#5e697b")

sim_ts <- map(sim_ts_, ~ predict(fit, .)) |>
  map_dfr(pivot_ts, .id = "counterfactual")

sim_ts |>
  filter(taxon %in% c("X_0001", "X_0019", "X_0129", "X_0009", "X_0012", "X_0033"), time > 20, time < 70) |>
  mutate(
    pH = factor(round(plateaus[as.integer(counterfactual)], 3)),
    taxon = factor(taxon, levels = taxa(ts)[taxa_order])
    ) |>
  group_by(taxon, time, pH) |>
  summarise(q50 = median(value)) |>
  ggplot() +
   geom_line(aes(time, q50, col = pH)) +
   geom_vline(col = "#c1c1c1c1", xintercept = L, linewidth = 1) +
   facet_wrap(~ taxon, scales = "free_y") +
   scale_fill_manual(values = pal) +
   scale_color_manual(values = pal)
```
