---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/",
  out.width = "100%"
)
```

# Microbiome Intervention Analysis using `mbtransfer`

<!-- badges: start -->
[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/krisrs1128/mbtransfer_demo/HEAD?urlpath=rstudio)
<!-- badges: end -->

`mbtransfer` is an R package for modeling and inference of microbial communities
under dynamic environmental interventions. It supports simulation of
hypothetical community trajectories under user-specified perturbations and can
be used to select for taxa that are significantly impacted by this change in a
statistically rigorous sense. It accomplishes this by learning a separate
transfer function model for each taxon, relating its changes in abundance to its
immediate community and environmental context. The fitted models can be used to
measure the significance of the intervention on individual taxa,

<center>
<img src=""/>
</center>

(larger mirror statistics correspond to stronger effects) and to simulate
trajectory differences between counterfactual treatments,

<center>
<img src=""/>
</center>

You can read more about the methodology in our preprint and can browse complete
examples in the [package documentation](https://krisrs1128.github.io/mbtransfer).

## Installation

You can install `mbtransfer` from this repository using:

```{r, eval = FALSE}
#install.packages("devtools")
devtools::install_github("krisrs1128/mbtransfer")
```

## Help

We welcome questions and comments about the package! You can reach us either
through [github](https://github.com/krisrs1128/mbtransfer/issues) or [email](https://krisrs1128.github.io/LSLab/_includes/contact).
